\bf Quantile Trend Filtering via Spingarn's Method

### 2018-01-10 Manuscript/detrendify_JCGS.tex

%
% A simple LaTeX template for Books
%  (c) Aleksander Morgado <aleksander@es.gnu.org>
%  Released into public domain
%

\documentclass{article}
\usepackage[a4paper, top=3cm, bottom=3cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage{setspace}
\usepackage{fancyhdr}
\usepackage{tocloft}
\usepackage[usenames]{color}
\RequirePackage[OT1]{fontenc}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\graphicspath{{./Figures/}}

%% For special lists like inparaenum, compactenum, compactitem
\usepackage{paralist}
%% For \boldsymbol
\usepackage{amsbsy}

@@

%% For \bm (bold math)
\usepackage{bm}

@@

@@

%% Algorithm
\usepackage{algorithm,algorithmic}
%\usepackage{algpseudocode}

%% Useful shorthand for converting equations to be inline
\newenvironment{inlinemath}{$}{$}

%% Define hypen
\mathchardef\mhyphen="2D

@@

@@

\ifpdf
\newcommand{\Sec}[1]{\hyperref[sec:#1]{\S\ref*{sec:#1}}} %section
\newcommand{\App}[1]{\hyperref[sec:#1]{Appendix~\ref*{sec:#1}}} %appendix
\newcommand{\Eqn}[1]{\hyperref[eq:#1]{{\rm (\ref*{eq:#1})}}} %equation
\newcommand{\Part}[1]{\hyperref[part:#1]{(\ref*{part:#1})}} %part of theorem
\newcommand{\Fig}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}} %figure
\newcommand{\Tab}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}} %table
\newcommand{\Thm}[1]{\hyperref[thm:#1]{Theorem~\ref*{thm:#1}}} %theorem
\newcommand{\Lem}[1]{\hyperref[lem:#1]{Lemma~\ref*{lem:#1}}} %lemma
\newcommand{\Prop}[1]{\hyperref[prop:#1]{Proposition~\ref*{prop:#1}}} %proposition
\newcommand{\Cor}[1]{\hyperref[cor:#1]{Corollary~\ref*{cor:#1}}} %corollary
\newcommand{\Def}[1]{\hyperref[def:#1]{Definition~\ref*{def:#1}}} %definition
\newcommand{\Alg}[1]{\hyperref[alg:#1]{Algorithm~\ref*{alg:#1}}} %algorithm
\newcommand{\Ex}[1]{\hyperref[ex:#1]{Example~\ref*{ex:#1}}} %example
\newcommand{\As}[1]{\hyperref[as:#1]{Assumption~{\rm\ref*{as:#1}}}} %assumption
\newcommand{\Reg}[1]{\hyperref[as:#1]{Condition~\ref*{reg:#1}}} %regularity condition
\newcommand{\AlgLine}[2]{\hyperref[alg:#1]{line~\ref*{line:#2} of Algorithm~\ref*{alg:#1}}}
\newcommand{\AlgLines}[3]{\hyperref[alg:#1]{lines~\ref*{line:#2}--\ref*{line:#3} of Algorithm~\ref*{alg:#1}}}
\else
\newcommand{\Sec}[1]{{\S\ref{sec:#1}}} %section
\newcommand{\App}[1]{{Appendix~\ref{sec:#1}}} %appendix
\newcommand{\Eqn}[1]{{(\ref{eq:#1})}} %equation
\newcommand{\Part}[1]{{(\ref{part:#1})}} %part of proof
\newcommand{\Fig}[1]{{Figure~\ref{fig:#1}}} %figure
\newcommand{\Tab}[1]{{Table~\ref{tab:#1}}} %table
\newcommand{\Thm}[1]{{Theorem~\ref{thm:#1}}} %theorem
\newcommand{\Lem}[1]{{Lemma~\ref{lem:#1}}} %lemma
\newcommand{\Prop}[1]{{Proposition~\ref{prop:#1}}} %proposition
\newcommand{\Cor}[1]{{Corollary~\ref{cor:#1}}} %corollary
\newcommand{\Def}[1]{{Definition~\ref{def:#1}}} %definition
\newcommand{\Alg}[1]{{Algorithm~\ref{alg:#1}}} %algorithm
\newcommand{\Ex}[1]{{Example~\ref{ex:#1}}} %example
\newcommand{\Reg}[1]{{R~\ref*{reg:#1}}} %regularity condition
\fi

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
%% Assumptions environment
\newtheorem{assumption}[theorem]{Assumption}

\begin{document}
\title{Quantile Trend Filtering via Spingarn's Method}
\noindent
\textbf{Quantile Trend Filtering via Spingarn's Method}\\
\textbf{Date: \today}\\

%% ----------------------------------------------------------------------
%% Introduction
%% ----------------------------------------------------------------------
\section{Introduction}

@@

\textbf{Oh, Lee, and Nychka 2011 - Fast Nonparametric Quantile Regression With Arbitrary Smoothing Method} Unified framework for non-parametric quantile regression by approximating check loss function with quadratic loss near zero.\\
\textbf{Kim, Koh, Boyd, and Gorinevsky 2009} - Introduce the concept of l1 trend filtering.\\
\textbf{Tibshirani, 2014 - Adaptive piecewise polynomial estimation via trend filtering} Describes properties of trend filtering using quadratic loss. Shows that trend filtering estimates adapt to the local level of smoothness much better than smoothing splines, and exhibit
a remarkable similarity to locally adaptive regression splines. Prove that (with the right choice of tuning parameter) the trend filtering estimate
converges to the true underlying function at the minimax rate for functions whose kth derivative is of bounded variation.\\
\textbf{Ning, Selesnick, Duval 2014 - Chromatogram baseline estimation and denoising using sparsity (BEADS)} Address problem of estimating a smooth baseline in noisy data with drift. \\
We propose to use the trend filtering penalty with the check loss function to produce a non-parametric quantile regression estimate that can be computes using a linear time algorithm for removing trends in time series. Can be extended to model multiple quantiles and ensure non-crossing.

@@

	 \begin{figure}[!h] 
	 	\caption{Raw Data - three collocated sensors}
		\includegraphics[width = 5in]{spodRaw.png}
	\end{figure}
	 \begin{figure}[!h] 
	 	\caption{Detrended data - three collocated sensors}
	\includegraphics[width = 5in]{spodDetrended.png}
	\end{figure}

Examples:
\begin{itemize}
\item Air quality \cite{Apte2017}
\item ECG \cite{Sanyal2012}
\item Chromatogram baseline estimation  \cite{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \cite{Ilewicz2015,Bacher2016}
\item Identifying absorption dips from black body radiation
\end{itemize}

Need to decide between detrending versus detrending + denoising. The BEADS \cite{Ning2014} method does both. We may wish to focus on detrending and then use wavelet SURE denoising as a postprocessing step, i.e. do a two-stage procedure, both of which can be done in linear time. On paper this should be faster than the BEADS procedure. Or we may just want to stick with detrending.  There's Matlab code for BEADS, and the BEADS paper also points to two other popular methods in chromotography.

Things to do:
\begin{itemize}
	\item Convergence of the algorithm
	\item Convergence rate \cite{He2012, He2015, Davis2017}
	\item Timing experiments of LP versus Spingarn
	\item Make an R package - detrendr
	\item Compare on synthetic data quality of solution with existing methods
	\item Do comparisons on real data examples
\end{itemize}

@@

The classic least squares regression is notoriously sensitive to outliers. One remedy to blunt the influence of outliers is to compute the least absolute deviations (LAD) solution in place of the least squares one. Given a design matrix $\M{X} \in \Real^{n \times p}$ and continuous responses $\V{y} \in \Real^n$, we estimate a regression vector $\V{\theta} \in \Real^p$ so that $\M{X}\V{\theta}$ is a good approximation of $\V{y}$. The LAD estimator is a solution to the problem
\begin{eqnarray*}
\underset{\V{\theta}}{\min}\; \frac{1}{n}\lVert \V{y} - \M{X}\V{\theta} \rVert_1.
\end{eqnarray*}
The above optimization problem generalizes the notion of the median of a collection of numbers. A median $\mu$ of $n$ reals $\VE{y}{1}, \ldots, \VE{y}{n}$ is the minimizer of the function
\begin{eqnarray*}
f(u) & = & \frac{1}{n}\sum_{i=1}^n \lvert \VE{y}{i} - \theta \rvert.
\end{eqnarray*}
Recall that the median is the 50th percentile or 0.5-quantile, namely half of the $\VE{y}{i}$ are less than or equal to $\mu$ and the other half is greater than or equal to $\mu$. The median can be generalized to arbitrary $\tau$-quantiles for $\tau \in (0,1)$ to give us quantile regression \cite{Koenker1978}.

@@

In the trend filtering problem \cite{Kim2009, Tib2014}, one is interested in finding an adaptive polynomial approximation to noisy data $\V{y} \in \Real^n$ by solving the following convex problem.
\begin{eqnarray*}
\underset{\V{\theta}}{\arg\min}\; \frac{1}{2n} \lVert \V{y} - \V{\theta} \rVert_2^2 + \lambda \lVert \Mn{D}{k+1}\V{\theta} \rVert_1,
\end{eqnarray*}
where $\lambda \geq 0$ is a regularization parameter that trades off the emphasis on the data fidelity term and the matrix $\Mn{D}{k+1} \in \Real^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\Mn{D}{k+1}$ consider the difference operator when $k = 0$.
\begin{eqnarray*}
\Mn{D}{1} = \begin{pmatrix}
-1 & 1 & 0 & \cdots & 0 & 0 \\
0 & -1 & 1 & \cdots & 0 & 0 \\
\vdots & & & & & \\
0 & 0 & 0 & \cdots & -1 & 1 \\
\end{pmatrix}
\end{eqnarray*}
Thus, $\lVert \Mn{D}{1}\V{\theta} \rVert_1 = \sum_{i=1}^{n-1} \lvert \VE{\theta}{i} - \VE{\theta}{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty incentivizes solutions which are piece-wise constant. For $k \geq 1$, the difference operator $\Mn{D}{k+1} \in \Real^{(n-k-1) \times n}$ is defined recursively as follows
\begin{eqnarray*}
\Mn{D}{k+1} & = & \Mn{D}{1}\Mn{D}{k}.
\end{eqnarray*}
By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$.

@@

\begin{itemize}
	\item Quantile splines \cite{Oh2011}
	\item BEADS 
\end{itemize}

@@

We first review Spingarn's method \cite{Spingarn1985}, which solves the following equality constrained convex problem:
\begin{equation}
\label{eq:spingarn}
\begin{split}
\text{minimize}\; & \psi(\V{x}) \\
\text{subject to}\; & \V{x} \in V,
\end{split}
\end{equation}
where $V$ is a subspace. The problem \Eqn{spingarn} can be expressed as the unconstrained optimization problem
\begin{equation}
\label{eq:spingarn2}
\text{minimize}\; \psi(\V{x}) + \iota_V(\V{x}),
\end{equation}
where $\iota_V$ is the indicator function of the set $V$. Spingarn's method applies Douglas-Rachford splitting to the problem \Eqn{spingarn2} to give the following updates.

\begin{eqnarray*}
\Vn{x}{k+1} & = & \prox{t\psi}{\Vn{z}{k}} \\
\Vn{y}{k+1} & = & P_V\left(2\Vn{x}{k+1} - \Vn{z}{k} \right) \\
\Vn{z}{k+1} & = & \Vn{z}{k} + \lambda^{(k)}\left(\Vn{y}{k+1} - \Vn{x}{k+1} \right).
\end{eqnarray*}
The parameter $t$ is a step-size and $\lambda^{(k)}$ is Krasnosel'ski{\u i}-Mann iteration (Need citation). We require that $\lambda^{(k)} \in ]0, 2[$ and that $\sum_n \lambda^{(k)} (2-\lambda^{(k)}) = \infty$. The mapping $P_V$ is the orthogonal projection onto the set $V$.  Note that the algorithm iterates $\Vn{x}{k}$ will converge to a solution to problem $\Eqn{spingarn}$ \cite{Combettes2005}. We need to experiment with different over / under-relaxation parameters $\lambda^{(k)}$ and step sizes $t$. It will converge if we take $\lambda^{(k)} = 1$ and $t = 1$, but we may be able to converge faster in practice by taking non-trivial values.

@@

If we set $\psi(\V{\theta}, \V{\eta}) = f_1(\V{\theta}) + f_2(\V{\eta})$ and $V = \{ \V{z}\Tra = (\V{\theta}\Tra, \V{\eta}\Tra) : \V{\eta} = \M{D}\V{\theta} \}$, then we can apply Spingarn's method. Note that
\begin{eqnarray*}
\prox{th}{\V{\theta}, \V{\eta}} & = & (\prox{tf_1}{\V{\theta}}, \prox{tf_2}{\V{\eta}}) \\
P_V(\V{\theta}, \V{\eta}) & = & \begin{pmatrix}
\M{I} \\ \M{D}
\end{pmatrix}
\begin{pmatrix}
\M{I} + \M{D}\Tra\M{D}
\end{pmatrix}\Inv
\begin{pmatrix}
\V{\theta} + \M{D}\Tra\V{\eta}
\end{pmatrix},
\end{eqnarray*}
where the projection $P_V$ requires a banded linear system solve, with bandwidth $k + 1$.
This linear solve can be accomplished in $\mathcal{O}(n(k + 1)^2)$. The first solve using a banded Cholesky decomposition requires $\mathcal{O}(n(k + 1)^2)$. Subsequent solves require
$\mathcal{O}(n(k+1))$. We can use RccpArmadillo to do banded Cholesky \cite{Eddelbuettel2014}.

@@

\begin{itemize}
\item The overall computational complexity is essentially linear $\mathcal{O}(nk^2)$ for the initial banded Cholesky decomposition and the per-iteration complexity is $\mathcal{O}(nk)$.
\item One could also apply Anderson acceleration \cite{Walker2011} to reduce the number of Spingarn updates, since the Douglas-Rachford algorithm is a fixed point algorithm.
\end{itemize}

@@

Following the specialized ADMM algorithm for trend filtering \cite{Ramdas2016}, we can also take advantage of fast exact solvers of the one-dimensional fussed lasso problem \cite{davies2001, Johnson2013}. The first method is based on taut strings, and the second is based on dynamic programming. Both of these methods run in linear time. A third exact method with worst case quadratic penalty but linear time in typical cases was proposed in \cite{Condat2013}. C code is available for all three methods. We should compare them.

@@

Difference with version 1: If we are interested in a 3rd order penalization, then $\M{D} = \Mn{D}{2}$ in version 2 as opposed to $\M{D} = \Mn{D}{3}$ in version 1. The proximal mapping for $f_2$ is solved using one of the exact solvers \cite{davies2001, Johnson2013, Condat2013}. Show some timing results between the different versions. Point readers to \cite{Ramdas2016} for a discussion of why this minor reparameterization may lead to speed up.

@@

We compare our detrending method with BEADS and the more general nonparametric quantile method introduced in  \cite{Oh2011}.

\bibliographystyle{abbrv}
\bibliography{detrendify}

>>>>

\documentclass[12pt]{article}
\RequirePackage{amsthm,amsmath,amsbsy,amsfonts}
\usepackage{bm,graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
%\RequirePackage{hypernat}
\usepackage{paralist, booktabs, multirow}
%\usepackage{url}

\newtheorem{lemma}{Lemma}
\newcommand{\blind}{0}
\newcommand{\trace}{\text{Trace}}
\numberwithin{equation}{section}
\theoremstyle{plain}
%\newtheorem{thm}{Theorem}[section]
%\newtheorem{mydef}{Definition}
%\newtheorem{remark}{Remark}

%% ----------------------------------------------------------------------
%% Packages
%% ----------------------------------------------------------------------
%\usepackage{amsmath,amsfonts,amssymb}

\usepackage{amssymb}
\usepackage[usenames]{color}
%\usepackage{amsbsy,amsthm,amsmath,amssymb}

@@

%% For \llbracket and \rrbracket, \varoast, \varoslash
\usepackage{stmaryrd}

%% For \boldsymbol
%\usepackage{amsbsy}

%%% For \bm (bold math)
%\usepackage{bm}

@@

%% For special lists like inparaenum, compactenum, compactitem
\usepackage{paralist}

%% Turn on/off options for PDF
\usepackage{ifpdf}

%% Graphics
\usepackage{graphicx}
\ifpdf
\usepackage{epstopdf}
% REMOVE NEXT LINE TO REVERT TO PNG FILES
\DeclareGraphicsExtensions{.eps,.pdf}
\else
\usepackage{epsfig}
\fi

@@

@@

%% For tabulary
\usepackage{tabulary}

%% Show labels
%\usepackage[notref,notcite]{showkeys}

%% Algorithm
\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\ifpdf
\newcommand{\Sec}[1]{\hyperref[sec:#1]{Section~\ref*{sec:#1}}} %section
\newcommand{\App}[1]{\hyperref[sec:#1]{Appendix~\ref*{sec:#1}}} %appendix
\newcommand{\Eqn}[1]{\hyperref[eq:#1]{{\rm (\ref*{eq:#1})}}} %equation
\newcommand{\Part}[1]{\hyperref[part:#1]{(\ref*{part:#1})}} %part of theorem
\newcommand{\Fig}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}} %figure
\newcommand{\Tab}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}} %table
\newcommand{\Thm}[1]{\hyperref[thm:#1]{Theorem~\ref*{thm:#1}}} %theorem
\newcommand{\Lem}[1]{\hyperref[lem:#1]{Lemma~\ref*{lem:#1}}} %lemma
\newcommand{\Prop}[1]{\hyperref[prop:#1]{Proposition~\ref*{prop:#1}}} %proposition
\newcommand{\Cor}[1]{\hyperref[cor:#1]{Corollary~\ref*{cor:#1}}} %corollary
\newcommand{\Def}[1]{\hyperref[def:#1]{Definition~\ref*{def:#1}}} %definition
\newcommand{\Alg}[1]{\hyperref[alg:#1]{Algorithm~\ref*{alg:#1}}} %algorithm
\newcommand{\Ex}[1]{\hyperref[ex:#1]{Example~\ref*{ex:#1}}} %example
\newcommand{\As}[1]{\hyperref[as:#1]{Assumption~{\rm\ref*{as:#1}}}} %assumption
\newcommand{\Reg}[1]{\hyperref[as:#1]{Condition~\ref*{reg:#1}}} %regularity condition
\newcommand{\AlgLine}[2]{\hyperref[alg:#1]{line~\ref*{line:#2} of Algorithm~\ref*{alg:#1}}}
\newcommand{\AlgLines}[3]{\hyperref[alg:#1]{lines~\ref*{line:#2}--\ref*{line:#3} of Algorithm~\ref*{alg:#1}}}
\else
\newcommand{\Sec}[1]{{Section~\ref{sec:#1}}} %section
\newcommand{\App}[1]{{Appendix~\ref{sec:#1}}} %appendix
\newcommand{\Eqn}[1]{{(\ref{eq:#1})}} %equation
\newcommand{\Part}[1]{{(\ref{part:#1})}} %part of proof
\newcommand{\Fig}[1]{{Figure~\ref{fig:#1}}} %figure
\newcommand{\Tab}[1]{{Table~\ref{tab:#1}}} %table
\newcommand{\Thm}[1]{{Theorem~\ref{thm:#1}}} %theorem
\newcommand{\Lem}[1]{{Lemma~\ref{lem:#1}}} %lemma
\newcommand{\Prop}[1]{{Property~\ref{prop:#1}}} %property
\newcommand{\Cor}[1]{{Corollary~\ref{cor:#1}}} %corollary
\newcommand{\Def}[1]{{Definition~\ref{def:#1}}} %definition
\newcommand{\Alg}[1]{{Algorithm~\ref{alg:#1}}} %algorithm
\newcommand{\Ex}[1]{{Example~\ref{ex:#1}}} %example
\newcommand{\As}[1]{{Assumption~\ref{as:#1}}} %assumption
\newcommand{\Reg}[1]{{R~\ref{reg:#1}}} %regularity condition
\newcommand{\AlgLine}[2]{{line~\ref{line:#2} of Algorithm~\ref{alg:#1}}}
\newcommand{\AlgLines}[3]{{lines~\ref{line:#2}--\ref{line:#3} of Algorithm~\ref{alg:#1}}}
\fi

\usepackage{paralist, booktabs, multirow}

%% Assumptions environments
\newtheorem{assumption}{Assumption}[section]
%% Special proof where box is inside final equation
% The function \proof and \endproof are defined inside the SIAM class file.
% Be sure to use \myproofend inside the final equation to produce the
% special box that signifies the end of the proof.
\newenvironment{myproof}{\proof}{}
\newcommand{\myproofend}{\quad\endproof}
\newtheorem{theorem}{Theorem}[section]

%Added by Arvind
\usepackage[normalem]{ulem}
\newcommand{\arvind}[1]{{\color{red} #1}}

%%%%%

%\RequirePackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
%
%\RequirePackage[OT1]{fontenc}
%\RequirePackage{amsthm,amsmath}
%\RequirePackage [round,authoryear]{natbib}
%\RequirePackage{hypernat}
%
%\usepackage{amsbsy,amsthm,amsmath,amssymb}
%\usepackage{graphicx,booktabs}
%%\usepackage{subfig}
\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usetikzlibrary{arrows}
%\setlength{\oddsidemargin}{.75in}
%\setlength{\evensidemargin}{.75in}
%\setlength{\textwidth}{5in}
\newtheorem{proposition}{Proposition}[section]
%\newtheorem{corollary}{Corollary}[section]
%\newtheorem{example}{Example}[section]
%\newtheorem{definition}{Definition}[section]

@@

%\startlocaldefs
%\numberwithin{equation}{section}
%\theoremstyle{plain}
%\newtheorem{thm}{Theorem}[section]
%\endlocaldefs

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\pdfminorversion=4

%% ----------------------------------------------------------------------
%% Main Document
%% ----------------------------------------------------------------------
\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Quantile Trend Filtering via Spingarn's Method}
  \author{Halley Brantley\thanks{
    Department of Statistics, North Carolina State University, Raleigh, NC 27695-8203 (E-mail: hlbrantl@ncsu.edu)} \,
         Joseph Guinness\thanks{
    Department of Statistics, North Carolina State University, Raleigh, NC 27695-8203 (E-mail: jsguinne@ncsu.edu)} \,
    and
%    and
    Eric C. Chi\thanks{Department of Statistics, North Carolina State University, Raleigh, NC 27695-8203 (E-mail: eric$\_$chi@ncsu.edu).}    \\}
    \date{}
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Abstract
\end{abstract}

\noindent%
{\it Keywords:}  Key words
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
%% ----------------------------------------------------------------------
%% Introduction
%% ----------------------------------------------------------------------
\section{Introduction}

@@

\begin{itemize}
\item \cite{Oh2011} unified framework for non-parametric quantile regression by approximating check loss function with quadratic loss near zero.
\item \cite{KoenkerNgPortnoy1994} introduce quantile smoothing splines. We need to discuss how our approach is different.
\item \cite{Kim2009} introduce the concept of $\ell_1$-trend filtering.
\item \cite{Tib2014} describes properties of trend filtering using quadratic loss. Shows that trend filtering estimates adapt to the local level of smoothness much better than smoothing splines, and exhibit a remarkable similarity to locally adaptive regression splines. Prove that (with the right choice of tuning parameter) the trend filtering estimate
converges to the true underlying function at the minimax rate for functions whose kth derivative is of bounded variation.
\item \cite{Ning2014} address problem of estimating a smooth baseline in noisy data with drift.
\end{itemize}
We propose to use the trend filtering penalty with the check loss function to produce a non-parametric quantile regression estimate that can be computes using a linear time algorithm for removing trends in time series. The formulation was proposed by \cite{Kim2009} as a possible extension of $\ell_1$-trend filtering but not studied. Moreover we extend the basic framework to model multiple quantiles and ensure non-crossing.

@@

	 \begin{figure}[!h] 
	 	\caption{Raw Data - three collocated sensors}
		\includegraphics[width = 5in]{Figures/spodRaw.png}
	\end{figure}
	 \begin{figure}[!h] 
	 	\caption{Detrended data - three collocated sensors}
	\includegraphics[width = 5in]{Figures/spodDetrended.png}
	\end{figure}

Examples:
\begin{itemize}
\item Air quality \citep{Apte2017}
\item ECG \citep{Sanyal2012}
\item Chromatogram baseline estimation  \citep{Ning2014, Ilewicz2015}
\item Galaxy spectrum baseline estimation \citep{Ilewicz2015,Bacher2016}
\item Identifying absorption dips from black body radiation
\end{itemize}

Need to decide between detrending versus detrending + denoising. The BEADS \citep{Ning2014} method does both. We may wish to focus on detrending and then use wavelet SURE denoising as a postprocessing step, i.e. do a two-stage procedure, both of which can be done in linear time. On paper this should be faster than the BEADS procedure. Or we may just want to stick with detrending.  There's Matlab code for BEADS, and the BEADS paper also points to two other popular methods in chromotography.

Things to do:
\begin{itemize}
	\item Convergence of the algorithm
	\item Convergence rate \citep{He2012, He2015, Davis2017}
	\item Timing experiments of LP versus Spingarn
	\item Make an R package - detrendr
	\item Compare on synthetic data quality of solution with existing methods
	\item Do comparisons on real data examples
\end{itemize}

@@

The classic least squares regression is notoriously sensitive to outliers. One remedy to blunt the influence of outliers is to compute the least absolute deviations (LAD) solution in place of the least squares one. Given a design matrix $\M{X} \in \Real^{n \times p}$ and continuous responses $\V{y} \in \Real^n$, we estimate a regression vector $\V{\theta} \in \Real^p$ so that $\M{X}\V{\theta}$ is a good approximation of $\V{y}$. The LAD estimator is a solution to the problem
\begin{eqnarray*}
\underset{\V{\theta}}{\min}\; \frac{1}{n}\lVert \V{y} - \M{X}\V{\theta} \rVert_1.
\end{eqnarray*}
The above optimization problem generalizes the notion of the median of a collection of numbers. A median $\mu$ of $n$ reals $\VE{y}{1}, \ldots, \VE{y}{n}$ is the minimizer of the function
\begin{eqnarray*}
f(u) & = & \frac{1}{n}\sum_{i=1}^n \lvert \VE{y}{i} - \theta \rvert.
\end{eqnarray*}
Recall that the median is the 50th percentile or 0.5-quantile, namely half of the $\VE{y}{i}$ are less than or equal to $\mu$ and the other half is greater than or equal to $\mu$. The median can be generalized to arbitrary $\tau$-quantiles for $\tau \in (0,1)$ to give us quantile regression \citep{Koenker1978}.

@@

In the trend filtering problem \citep{Kim2009, Tib2014}, one is interested in finding an adaptive polynomial approximation to noisy data $\V{y} \in \Real^n$ by solving the following convex problem.
\begin{eqnarray*}
\underset{\V{\theta}}{\arg\min}\; \frac{1}{2n} \lVert \V{y} - \V{\theta} \rVert_2^2 + \lambda \lVert \Mn{D}{k+1}\V{\theta} \rVert_1,
\end{eqnarray*}
where $\lambda \geq 0$ is a regularization parameter that trades off the emphasis on the data fidelity term and the matrix $\Mn{D}{k+1} \in \Real^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\Mn{D}{k+1}$ consider the difference operator when $k = 0$.
\begin{eqnarray*}
\Mn{D}{1} = \begin{pmatrix}
-1 & 1 & 0 & \cdots & 0 & 0 \\
0 & -1 & 1 & \cdots & 0 & 0 \\
\vdots & & & & & \\
0 & 0 & 0 & \cdots & -1 & 1 \\
\end{pmatrix}
\end{eqnarray*}
Thus, $\lVert \Mn{D}{1}\V{\theta} \rVert_1 = \sum_{i=1}^{n-1} \lvert \VE{\theta}{i} - \VE{\theta}{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty incentivizes solutions which are piece-wise constant. For $k \geq 1$, the difference operator $\Mn{D}{k+1} \in \Real^{(n-k-1) \times n}$ is defined recursively as follows
\begin{eqnarray*}
\Mn{D}{k+1} & = & \Mn{D}{1}\Mn{D}{k}.
\end{eqnarray*}
By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$.

@@

\begin{itemize}
	\item Quantile splines \citep{Oh2011}
	\item BEADS 
\end{itemize}

@@

We first review Spingarn's method \citep{Spingarn1985}, which solves the following equality constrained convex problem:
\begin{equation}
\label{eq:spingarn}
\begin{split}
\text{minimize}\; & \psi(\V{x}) \\
\text{subject to}\; & \V{x} \in V,
\end{split}
\end{equation}
where $V$ is a subspace. The problem \Eqn{spingarn} can be expressed as the unconstrained optimization problem
\begin{equation}
\label{eq:spingarn2}
\text{minimize}\; \psi(\V{x}) + \iota_V(\V{x}),
\end{equation}
where $\iota_V$ is the indicator function of the set $V$. Spingarn's method applies Douglas-Rachford splitting to the problem \Eqn{spingarn2} to give the following updates.

\begin{eqnarray*}
\Vn{x}{k+1} & = & \prox{t\psi}{\Vn{z}{k}} \\
\Vn{y}{k+1} & = & P_V\left(2\Vn{x}{k+1} - \Vn{z}{k} \right) \\
\Vn{z}{k+1} & = & \Vn{z}{k} + \lambda^{(k)}\left(\Vn{y}{k+1} - \Vn{x}{k+1} \right).
\end{eqnarray*}
The parameter $t$ is a step-size and $\lambda^{(k)}$ is Krasnosel'ski{\u i}-Mann iteration (Need citation). We require that $\lambda^{(k)} \in ]0, 2[$ and that $\sum_n \lambda^{(k)} (2-\lambda^{(k)}) = \infty$. The mapping $P_V$ is the orthogonal projection onto the set $V$.  Note that the algorithm iterates $\Vn{x}{k}$ will converge to a solution to problem $\Eqn{spingarn}$ \citep{Combettes2005}. We need to experiment with different over / under-relaxation parameters $\lambda^{(k)}$ and step sizes $t$. It will converge if we take $\lambda^{(k)} = 1$ and $t = 1$, but we may be able to converge faster in practice by taking non-trivial values.

@@

If we set $\psi(\V{\theta}, \V{\eta}) = f_1(\V{\theta}) + f_2(\V{\eta})$ and $V = \{ \V{z}\Tra = (\V{\theta}\Tra, \V{\eta}\Tra) : \V{\eta} = \M{D}\V{\theta} \}$, then we can apply Spingarn's method. Note that
\begin{eqnarray*}
\prox{th}{\V{\theta}, \V{\eta}} & = & (\prox{tf_1}{\V{\theta}}, \prox{tf_2}{\V{\eta}}) \\
P_V(\V{\theta}, \V{\eta}) & = & \begin{pmatrix}
\M{I} \\ \M{D}
\end{pmatrix}
\begin{pmatrix}
\M{I} + \M{D}\Tra\M{D}
\end{pmatrix}\Inv
\begin{pmatrix}
\V{\theta} + \M{D}\Tra\V{\eta}
\end{pmatrix},
\end{eqnarray*}
where the projection $P_V$ requires a banded linear system solve, with bandwidth $k + 1$.
This linear solve can be accomplished in $\mathcal{O}(n(k + 1)^2)$. The first solve using a banded Cholesky decomposition requires $\mathcal{O}(n(k + 1)^2)$. Subsequent solves require
$\mathcal{O}(n(k+1))$. We can use RccpArmadillo to do banded Cholesky \citep{Eddelbuettel2014}.

@@

\begin{itemize}
\item The overall computational complexity is essentially linear $\mathcal{O}(nk^2)$ for the initial banded Cholesky decomposition and the per-iteration complexity is $\mathcal{O}(nk)$.
\item One could also apply Anderson acceleration \citep{Walker2011} to reduce the number of Spingarn updates, since the Douglas-Rachford algorithm is a fixed point algorithm.
\end{itemize}

@@

Following the specialized ADMM algorithm for trend filtering \citep{Ramdas2016}, we can also take advantage of fast exact solvers of the one-dimensional fussed lasso problem \citep{davies2001, Johnson2013}. The first method is based on taut strings, and the second is based on dynamic programming. Both of these methods run in linear time. A third exact method with worst case quadratic penalty but linear time in typical cases was proposed by \cite{Condat2013}. C code is available for all three methods. We should compare them.

@@

Difference with version 1: If we are interested in a 3rd order penalization, then $\M{D} = \Mn{D}{2}$ in version 2 as opposed to $\M{D} = \Mn{D}{3}$ in version 1. The proximal mapping for $f_2$ is solved using one of the exact solvers \citep{davies2001, Johnson2013, Condat2013}. Show some timing results between the different versions. Point readers to \cite{Ramdas2016} for a discussion of why this minor reparameterization may lead to speed up.

@@

We compare our detrending method with BEADS and the more general nonparametric quantile method introduced by \cite{Oh2011}.

\bibliographystyle{asa}
\bibliography{detrendify}
